

## load data from GEO database - GSE249721

```{r}
library(GEOquery)
```

## get the datat from the supplementary files
```{r}
# download the supplementary files and save them in a specific folder
path_save <- "/Users/alvise/Documents/uni/semestre3/laboratory_bio_datamining/data/RNA/expression_matrices"
supp_files <- getGEOSuppFiles("GSE249721", baseDir = path_save)
# check the downloaded files
list.files(path_save)
# get the file name
file_path <- rownames(supp_files)[1]
# Load the full file, skipping the '#' comment lines
full_data <- read.delim(
  file = file_path, 
  header = TRUE, 
  comment.char = "#"
)
head(full_data)
```

save data frame as RDS file for future use
```{r}
path_1 <- "/Users/alvise/Documents/uni/semestre3/laboratory_bio_datamining/data/RNA/expression_matrices/GSE249721"
matrix_path <- file.path(path_1, "GSE249721_full_data.rds")
saveRDS(full_data, file = matrix_path)
```

load the RDS file
```{r}
matrix_path <- "/Users/alvise/Documents/uni/semestre3/laboratory_bio_datamining/data/RNA/expression_matrices/GSE249721/GSE249721_full_data.rds"
full_data <- readRDS(file = matrix_path)
head(full_data) 
dim(full_data)  
````

### extract the expression matrix of the HCC827 cell line
```{r}
# subset the data for HCC827 samples
hcc827_data <- full_data[, grepl("HCC827", colnames(full_data))]
# check the dimensions of the subsetted data
dim(hcc827_data)
print(hcc827_data[1:6, 1:6])
```

## preprocessing the data

create groups based on treatment conditions
the two groups are: "Osimertinib" and "Untreated"
```{r}
# create a new data frame with the sample names and their corresponding groups
sample_info <- data.frame(
  Sample = colnames(hcc827_data),
  Group = ifelse(grepl("Osimertinib", colnames(hcc827_data)), "Osimertinib", "Untreated")
)

# check the sample information
head(sample_info)
```

 create a box plot to visualize the distribution of expression values across samples, include the two groups with different colors
```{r}
png(filename = "hcc827_expression_boxplot.png", width = 800, height = 600)
boxplot(hcc827_data, 
        main = "Boxplot of HCC827 Expression Data",
        xlab = "Samples",
        ylab = "Expression Values",
        las = 2, # rotate x-axis labels for better readability
        col = ifelse(sample_info$Group == "Osimertinib", "lightblue", "lightgreen"))
        legend("topright", legend = c("Osimertinib", "Untreated"), fill = c("lightblue", "lightgreen"))
dev.off()

```

as the expression values are between 0 and 15 we can asume the data has been alread log2 transformed 

```{r}
library(limma)

# --- NEW FILTERING STEP ---
# 1. Calculate the variance for every gene (row)
gene_variances <- apply(hcc827_data, 1, var)

# 2. Keep only genes that have a variance greater than 0
# (You could even use a higher threshold, like 0.1, to be stricter)
keep <- gene_variances > 0
filtered_data <- hcc827_data[keep, ]

# 3. See how many genes we kept
print(paste("Original genes:", nrow(hcc827_data)))
print(paste("Filtered genes:", nrow(filtered_data)))
# --- END OF FILTERING ---


# 4. Create the design matrix (same as before)
design <- model.matrix(~ 0 + Group, data = sample_info)
colnames(design) <- c("Osimertinib", "Untreated")
contrast.matrix <- makeContrasts(Osimertinib - Untreated, levels = design)

# 5. Fit the model *on the filtered data*
fit <- lmFit(filtered_data, design) # <-- Use filtered_data
fit2 <- contrasts.fit(fit, contrast.matrix)
fit2 <- eBayes(fit2)

# 6. Get the results (which should now be much more reasonable)
top_genes <- topTable(fit2, n = Inf)

# 7. Check the new, more realistic results
print(head(top_genes))

# 8. NOW, look for your gene
# (Replace "THE_REAL_GENE_NAME" with what you found in the grep step)
# "THE_REAL_GENE_NAME" %in% rownames(top_genes)
# check the gene names
print(rownames(top_genes)[1:20])
# the gene names are numbers of 4 digits
# ------------------------------
# ------------------------------
# Load the annotation library
library(org.Hs.eg.db)

# 1. Get the Entrez IDs from your results
entrez_ids <- rownames(top_genes)

# 2. Use the 'mapIds' function to translate them
# We are telling it: "Here are Entrez IDs, please give me the SYMBOLs"
gene_symbols <- mapIds(org.Hs.eg.db,
                       keys = entrez_ids,
                       column = "SYMBOL",
                       keytype = "ENTREZID",
                       multiVals = "first")

# 3. Add these new symbols as a column to your results
top_genes$symbol <- gene_symbols

# 4. Check the results!
print(head(top_genes, 20))
```


focus on finding the id for EGFR gene and its expression results

```{r}
library(org.Hs.eg.db)

egfr_id <- mapIds(org.Hs.eg.db,
                  keys = "EGFR",
                  column = "ENTREZID",
                  keytype = "SYMBOL")

print(egfr_id)
# check if EGFR is in the results
"1910" %in% rownames(top_genes) # -> TRUE
# extract the EGFR row
egfr_results <- top_genes["1910", ]
print(egfr_results)
```