## get the datat from the supplementary files

```{r}
library(GEOquery)
# download the supplementary files and save them in a specific folder
path_save <- "/Users/alvise/Documents/uni/semestre3/laboratory_bio_datamining/data/RNA/expression_matrices"
supp_files <- getGEOSuppFiles("GSE274293", baseDir = path_save)
# check the downloaded files
list.files(path_save)
# get the file name
file_path <- "/Users/alvise/Documents/uni/semestre3/laboratory_bio_datamining/data/RNA/expression_matrices/GSE274293/GSE274293_TPM.tsv.gz"
# Load the full file, skipping the '#' comment lines
full_data <- read.delim(
  file = file_path, 
  header = TRUE, 
  comment.char = "#"
)
head(full_data)
```

save file as RDS for future use
```{r}
path_1 <- "/Users/alvise/Documents/uni/semestre3/laboratory_bio_datamining/data/RNA/expression_matrices/GSE274293"
matrix_path <- file.path(path_1, "GSE274293_full_data.rds")
saveRDS(full_data, file = matrix_path)
```

## load the RDS file
```{r}
matrix_path <- "/Users/alvise/Documents/uni/semestre3/laboratory_bio_datamining/data/RNA/expression_matrices/GSE274293/GSE274293_full_data.rds"
full_data <- readRDS(file = matrix_path)
head(full_data)
dim(full_data)
```

## preprocessing the data (if needed)

create groups bases on treatment conditions: in this case it is case 1 and 2 in pre and post treatment. 
the two groups are : "Pre-treatment" and "Post-treatment"
```{r}
# create a new data frame with the sample names and their corresponding groups
sample_info <- data.frame(
  Sample = colnames(full_data)[-1], # exclude the first column if it contains gene identifiers
  Group = ifelse(grepl("post", colnames(full_data)[-1]), "post_treatment", "pre_treatment")
)   
# check the sample information
head(sample_info)
```

create a boxplot to visualize the distribution of expression values across samples
```{r}
png("/Users/alvise/Documents/uni/semestre3/laboratory_bio_datamining/assessing_mutation_frequency_EGFR_NSCLC/RNA_data_analysis/images/boxplot_GSE274293.png", width = 800, height = 600)
boxplot(full_data[,-1], # exclude the first column if it contains gene identifiers
        las = 2,
        main = "Expression values across samples in GSE274293",
        ylab = "Expression values (TPM)",
        col = ifelse(sample_info$Group == "Pre-treatment", "lightblue", "lightgreen"))
dev.off()

```

the boxplot shows the distribution of expression values across the samples. the y axis shows values
of TPM that go from 0 to 50000. this indicates that the data is not log transformed.
therefore, a log2 transformation is needed to normalize the data.

log 2 transformation of the data
```{r}
# apply log2 transformation to the expression data (excluding the first column if it contains gene identifiers)
log2_data <- full_data
log2_data[,-1] <- log2(log2_data[,-1] + 1) # adding 1 to avoid log(0)   
# check the first few rows of the log2 transformed data
head(log2_data)
```

do the boxplot again after log2 transformation
```{r}
png("/Users/alvise/Documents/uni/semestre3/laboratory_bio_datamining/assessing_mutation_frequency_EGFR_NSCLC/RNA_data_analysis/images/boxplot_GSE274293_log2.png", width = 800, height = 600)
boxplot(log2_data[,-1], # exclude the first column if it contains gene identifiers
        las = 2,
        main = "Log2 Transformed Expression values across samples in GSE274293",
        ylab = "Log2 Expression values (TPM)",
        col = ifelse(sample_info$Group == "Pre-treatment", "lightblue", "lightgreen"))
dev.off()
```

the median expression values are somewhat comparable. as the file from the dataset says TPM, I assume that the data is already normalized.

as the data is already normalized in TPM by the autors and i have log 2 transformed it. it is not possible to use deseq2 or edger as 
they need raw counts as imput. therefore, i will use limma for the differential expression analysis.

## differential expression analysis using limma

take into account that the data are paired samples (pre and post treatment from the same patient)

For this reason the samples are not independent. 
```{r}
library(limma)
# --- 1. Fix Your Data Format ---
# (Assuming 'log2_data' has gene IDs in column 1)
# Store the gene names
gene_ids <- log2_data[, 1]
# Create the purely numeric matrix
log2_matrix <- as.matrix(log2_data[, -1])
# Set the rownames
rownames(log2_matrix) <- gene_ids
# head(log2_matrix)

# --- 2. Create the Correct Paired 'sample_info' ---
# (Re-create this from your sample names)
sample_info_2 <- data.frame(
  row.names = colnames(log2_matrix),
  Patient = c("Case1", "Case1", "Case2", "Case2"), # Or whatever patient IDs
  Condition = c("Pre", "Post", "Pre", "Post")
)

# Set "Pre" as the baseline
sample_info_2$Condition <- factor(sample_info_2$Condition, levels = c("Pre", "Post"))
sample_info_2$Patient <- factor(sample_info_2$Patient)
# head(sample_info_2)

# --- 2.1 NEW: Filter the Matrix ---

# Filter 1: Remove genes with zero variance (This fixes your warning)
variances <- apply(log2_matrix, 1, var)
keep1 <- variances > 0

# Filter 2: (Better) Keep genes expressed in at least 2 samples
# We'll define "expressed" as a log2(TPM+1) value > 1 (i.e., TPM > 1)
keep2 <- rowSums(log2_matrix > 1) >= 2

# Combine the filters:
keep_final <- keep1 & keep2

# Create your final, filtered matrix
filtered_log2_matrix <- log2_matrix[keep_final, ]

print(paste("Original genes:", nrow(log2_matrix)))
print(paste("Filtered genes:", nrow(filtered_log2_matrix)))


# --- 3. Build the Correct Paired Model ---
# This model finds the 'Condition' effect *after* blocking for each 'Patient'
design <- model.matrix(~ Patient + Condition, data = sample_info_2)

# --- 4. Run the limma Pipeline ---
# Fit the linear model
fit <- lmFit(filtered_log2_matrix, design)

# Apply empirical Bayes moderation
fit2 <- eBayes(fit)

# Get top genes for the 'ConditionPost' effect
# This is the "Post vs. Pre" comparison
top_genes <- topTable(fit2, coef = "ConditionPost", number = Inf, adjust.method = "BH")

# View the (now correct) top differentially expressed genes
head(top_genes)
```

now i obtained a list of differentially expressed genes between pre and post treatment samples, taking into account the paired nature of the data.
I considered the global variance of the genes after filtering out the genes with low variance and low expression across samples. 
  This is done with the bayes moderation step of limma. And helps to obtain more reliable results, especially with small sample sizes such as in this case. 

## convert the gene identifiers to gene symbols

```{r}

# 1. Install/load the human genome annotation database
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
# BiocManager::install("org.Hs.eg.db")
library(org.Hs.eg.db)

# 2. Get the Ensembl IDs from your results (as a character vector)
ensembl_ids_with_version <- as.character(rownames(top_genes))

# --- 3. NEW STEP: Remove version numbers ---
# This uses a regular expression to find the literal dot ("\\.") 
# and any character after it (".*"), and replaces it with nothing ("")
ensembl_ids_cleaned <- sub("\\..*", "", ensembl_ids_with_version)

# 4. Use mapIds to translate the *cleaned* IDs. 
gene_symbols_map <- mapIds(org.Hs.eg.db,
                           keys = ensembl_ids_cleaned, # Use the cleaned IDs
                           column = "SYMBOL",
                           keytype = "ENSEMBL",
                           multiVals = "first")

# 5. Add symbols as a new column
# This works by looking up each 'cleaned_id' in the 'gene_symbols_map'
# and assigning the corresponding symbol to the correct row.
top_genes$symbol <- gene_symbols_map[ensembl_ids_cleaned]

# 6. Check your new, more readable table!
print(head(top_genes, 40))
```


## now check the presence of EGFR in the results
```{r}
# Check if "EGFR" is present, ignoring NA values
present <- "EGFR" %in% top_genes$symbol
print(paste("Is EGFR present in the results?", present))

# If present, show the EGFR row
if (present) {
  # Use which() to get the row index, this safely ignores NAs
  egfr_row <- top_genes[which(top_genes$symbol == "EGFR"), ]
  
  # A simpler way is to use the subset() function
  # egfr_row <- subset(top_genes, symbol == "EGFR") 
  
  print("Data for EGFR:")
  print(egfr_row)
} else {
  print("EGFR was not found in the significant results table.")
}
```

can i print the expression values of EGFR in pre and post treatment samples?
```{r}
# 1. Get the Ensembl ID for EGFR (which we know is "ENSG00000146648")
egfr_id <- "ENSG00000146648"

# 2. Check if this ID is in your filtered matrix rownames
egfr_present <- egfr_id %in% rownames(filtered_log2_matrix)

print(paste("Is EGFR in the filtered matrix?", egfr_present))

# 3. If it's there, print the expression values
if (egfr_present) {
  print("Log2(TPM+1) expression values for EGFR:")
  
  # Extract the row from your filtered_log2_matrix
  egfr_values <- filtered_log2_matrix[egfr_id, ]
  
  print(egfr_values)
}
```
from the output i can see that in case 1 the expression of EGFR is stable while in case 2 there is a decrease in expression after the treatment 
from 5.49 to 2.26 log2(TPM+1).

"Log2(TPM+1) expression values for EGFR:"
 Case1_pre Case1_post  Case2_pre Case2_post 
  2.348546   2.381839   5.497086   2.264044 

## construct a heatmap of the top 50 differentially expressed genes and include EGFR

```{r}
# 1. Install/load the pheatmap package
if (!require("pheatmap", quietly = TRUE))
    install.packages("pheatmap")
library(pheatmap)

# --- MODIFICATIONS START HERE ---

# 2. Get the Ensembl IDs of your top 50 genes
top_50_ids <- rownames(top_genes)[1:50]

# 2b. Define the Ensembl ID for EGFR (from your previous step)
egfr_id <- "ENSG00000146648"

# 2c. Create your final list of genes to plot
# union() adds EGFR and ensures there are no duplicates if it was already in the top 50
genes_to_plot_ids <- union(top_50_ids, egfr_id)

# --- MODIFICATIONS END HERE ---


# 3. Get the expression data *for this new list* from your filtered matrix
heatmap_data <- filtered_log2_matrix[genes_to_plot_ids, ]

# 4. Get the gene symbols for these genes
heatmap_labels <- top_genes[genes_to_plot_ids, "symbol"]

# 5. Set the row names of our data to be the symbols
# (This is a safer way to set labels, in case any symbols are NA)
final_labels <- ifelse(is.na(heatmap_labels), genes_to_plot_ids, heatmap_labels)
rownames(heatmap_data) <- final_labels

# 6. Get the sample annotation (from sample_info_2)
annotation_df <- data.frame(
  Condition = sample_info_2$Condition,
  Patient = sample_info_2$Patient,
  row.names = rownames(sample_info_2)
)

# 7. Create the heatmap!
pheatmap(
  heatmap_data,
  annotation_col = annotation_df,  # Adds colored bars for Patient/Condition
  scale = "row",                  # Z-scores each gene (CRITICAL for visualization)
  cluster_cols = FALSE,           # Don't cluster columns (keeps pre/post together)
  main = "Top 50 DEGs + EGFR (Post vs. Pre)" # <-- Updated title
)
```


## construct a volcanoplot (new style) and point out EGFR
```{r}
# --- 1. Load required libraries ---
library(ggplot2)
library(ggrepel)
library(dplyr) # For the 'mutate' function

# --- 2. Prepare the data ---
# Your data is 'top_genes'. We'll use this.
res_df <- top_genes

# --- 3. THIS IS THE KEY: Create the "class" column ---
# We define the thresholds from your example code
lfc_threshold <- 1.5
pval_threshold <- 0.05

res_df <- res_df %>%
  mutate(
    # Create the 'class' column based on logFC and P.Value
    class = case_when(
      logFC > lfc_threshold & P.Value < pval_threshold ~ "+",
      logFC < -lfc_threshold & P.Value < pval_threshold ~ "-",
      TRUE ~ "="
    ),
    # Make 'class' a factor to control the order in the legend
    class = factor(class, levels = c("+", "-", "="))
  )

# --- 4. Create the label for EGFR ---
res_df$gene_label <- ifelse(res_df$symbol == "EGFR", "EGFR", NA)

# --- 5. Create the plot (using the "king" style) ---
my_volcano_plot_gse274293 <- ggplot(res_df, aes(x = logFC, y = -log10(P.Value))) +
  
  # Draw all points, colored by our new 'class' column
  geom_point(aes(color = class), alpha = 0.7) +
  
  # Add the text label for EGFR (it will be black by default)
  geom_text_repel(aes(label = gene_label), na.rm = TRUE, box.padding = 0.5) +
  
  # Apply the colors from your example code
  scale_color_manual(values = c("=" = "grey70", "+" = "firebrick3", "-" = "blue3")) +
  
  # Add the dashed threshold lines
  geom_vline(xintercept = c(-lfc_threshold, lfc_threshold), linetype = "dashed", color = "grey20") +
  geom_hline(yintercept = -log10(pval_threshold), linetype = "dashed", color = "grey20") +
  
  # Add the theme
  theme_bw() +
  
  # Add labs
  labs(
    title = "Volcano Plot: Post vs. Pre-Treatment (GSE274293)",
    subtitle = "Significance: Uncorrected P-value < 0.05 and |logFC| > 1.5",
    x = "log2 Fold Change",
    y = "-log10 (Uncorrected p-value)",
    color = "Significance" # Adds a title to the legend
  )

# --- 6. Print the plot ---
print(my_volcano_plot_gse274293)

# --- 7. Save the plot as a square PNG file ---
ggsave(
  filename = "/Users/alvise/Documents/uni/semestre3/laboratory_bio_datamining/assessing_mutation_frequency_EGFR_NSCLC/RNA_data_analysis/images/GSE274293_volcano_plot_styled.png",
  plot = my_volcano_plot_gse274293,
  width = 8,
  height = 8,
  units = "in",
  dpi = 300
)
```


