

## load data from GEO database - GSE249721

```{r}
library(GEOquery)
```

## get the datat from the supplementary files
```{r}
# download the supplementary files and save them in a specific folder
path_save <- "/Users/alvise/Documents/uni/semestre3/laboratory_bio_datamining/data/RNA/expression_matrices"
supp_files <- getGEOSuppFiles("GSE249721", baseDir = path_save)
# check the downloaded files
list.files(path_save)
# get the file name
file_path <- rownames(supp_files)[1]
# Load the full file, skipping the '#' comment lines
full_data <- read.delim(
  file = file_path, 
  header = TRUE, 
  comment.char = "#"
)
head(full_data)
```

save data frame as RDS file for future use
```{r}
path_1 <- "/Users/alvise/Documents/uni/semestre3/laboratory_bio_datamining/data/RNA/expression_matrices/GSE249721"
matrix_path <- file.path(path_1, "GSE249721_full_data.rds")
saveRDS(full_data, file = matrix_path)
```

load the RDS file
```{r}
matrix_path <- "/Users/alvise/Documents/uni/semestre3/laboratory_bio_datamining/data/RNA/expression_matrices/GSE249721/GSE249721_full_data.rds"
full_data <- readRDS(file = matrix_path)
head(full_data) 
dim(full_data)  
````

### extract the expression matrix of the HCC827 cell line
```{r}
# subset the data for HCC827 samples
hcc827_data <- full_data[, grepl("HCC827", colnames(full_data))]
# check the dimensions of the subsetted data
dim(hcc827_data)
print(hcc827_data[1:6, 1:6])
```

## preprocessing the data

create groups based on treatment conditions
the two groups are: "Osimertinib" and "Untreated"
```{r}
# create a new data frame with the sample names and their corresponding groups
sample_info <- data.frame(
  Sample = colnames(hcc827_data),
  Group = ifelse(grepl("Osimertinib", colnames(hcc827_data)), "Osimertinib", "Untreated")
)

# check the sample information
head(sample_info)
```

 create a box plot to visualize the distribution of expression values across samples, include the two groups with different colors
save image file here: laboratory_bio_datamining/assessing_mutation_frequency_EGFR_NSCLC/RNA_data_analysis/images

```{r}
png(filename = "laboratory_bio_datamining/assessing_mutation_frequency_EGFR_NSCLC/RNA_data_analysis/images/hcc827_expression_boxplot_GSE249721.png", width = 800, height = 600) 
boxplot(hcc827_data, 
        main = "Boxplot of HCC827 Expression Data",
        xlab = "Samples",
        ylab = "Expression Values",
        las = 2, # rotate x-axis labels for better readability
        col = ifelse(sample_info$Group == "Osimertinib", "lightblue", "lightgreen"))
        legend("topright", legend = c("Osimertinib", "Untreated"), fill = c("lightblue", "lightgreen"))
dev.off()

```

as the expression values are between 0 and 15 we can asume the data has been alread log2 transformed 

## now we can proceed to differential expression analysis using limma package
```{r}
library(limma)

# --- NEW FILTERING STEP ---
# 1. Calculate the variance for every gene (row)
gene_variances <- apply(hcc827_data, 1, var)

# 2. Keep only genes that have a variance greater than 0
# (You could even use a higher threshold, like 0.1, to be stricter)
keep <- gene_variances > 0
filtered_data <- hcc827_data[keep, ]

# 3. See how many genes we kept
print(paste("Original genes:", nrow(hcc827_data)))
print(paste("Filtered genes:", nrow(filtered_data)))
# --- END OF FILTERING ---


# 4. Create the design matrix (same as before)
design <- model.matrix(~ 0 + Group, data = sample_info)
colnames(design) <- c("Osimertinib", "Untreated")
contrast.matrix <- makeContrasts(Osimertinib - Untreated, levels = design)

# 5. Fit the model *on the filtered data*
fit <- lmFit(filtered_data, design) # <-- Use filtered_data
fit2 <- contrasts.fit(fit, contrast.matrix)
fit2 <- eBayes(fit2)

# 6. Get the results (which should now be much more reasonable)
top_genes <- topTable(fit2, n = Inf)

# 7. Check the new, more realistic results
print(head(top_genes))
````

now we can proceed to translated the gene ids to gene symbols. we are proably dealing with Entrez IDs, but we need to check this first.
```{r}

# 8. NOW, look for your gene
# (Replace "THE_REAL_GENE_NAME" with what you found in the grep step)
# "THE_REAL_GENE_NAME" %in% rownames(top_genes)
# check the gene names
print(rownames(top_genes)[1:20])
# the gene names are numbers of 4 digits
# ------------------------------
# ------------------------------
# Load the annotation library
library(org.Hs.eg.db)

# 1. Get the Entrez IDs from your results
entrez_ids <- rownames(top_genes)

# 2. Use the 'mapIds' function to translate them
# We are telling it: "Here are Entrez IDs, please give me the SYMBOLs"
gene_symbols <- mapIds(org.Hs.eg.db,
                       keys = entrez_ids,
                       column = "SYMBOL",
                       keytype = "ENTREZID",
                       multiVals = "first")

# 3. Add these new symbols as a column to your results
top_genes$symbol <- gene_symbols

# ALTERNATIVE

# 2. Get the Entrez IDs from your results (e.g., "20799", "10502")
entrez_ids <- as.character(rownames(top_genes))

# 3. Use mapIds and specify keytype = "ENTREZID"
gene_symbol_2 <- mapIds(org.Hs.eg.db,
                       keys = entrez_ids,
                       column = "SYMBOL",
                       keytype = "ENTREZID", # <-- This is the correct keytype
                       multiVals = "first")

# 4. Add these symbols as a new column to your results table
top_genes$symbol_2 <- gene_symbol_2

# 5. Check your new, more readable table!
print(head(top_genes, 50))

# CHECK if the tow symbol columns are identical
all(top_genes$symbol == top_genes$symbol_2, na.rm = TRUE)
```


focus on finding the id for EGFR gene and its expression results

```{r}
library(org.Hs.eg.db)

egfr_id <- mapIds(org.Hs.eg.db,
                  keys = "EGFR",
                  column = "ENTREZID",
                  keytype = "SYMBOL")

print(egfr_id)
str(egfr_id) # should be a character vector of length 1
# check if EGFR is in the results
"1956" %in% rownames(top_genes) # -> TRUE
# extract the EGFR row
egfr_results <- top_genes["1956", ]
print(egfr_results)
# print 1956 row with gene symbol
egfr_results_with_symbol <- top_genes[rownames(top_genes) == "1956", ]
print(egfr_results_with_symbol)
```
EGFR differential expression results:
          logFC  AveExpr         t     P.Value   adj.P.Val         B symbol
  1956 -1.143036 10.40191 -7.398521 0.000250821 0.001646928 0.7437499   EGFR


## construct a volcano plot to visualize the differential expression results, highlighting the EGFR gene
```{r}

```


## construct a heatmap to visualize the expression of the top 50 differentially expressed genes, including EGFR

```{r}
library(pheatmap)
# select the top 50 differentially expressed genes
top_50_ids <- rownames(top_genes)[1:50]
# also include EGFR if not already in the top 50
if (!"1956" %in% top_50_ids) {
  top_50_ids <- c(top_50_ids, "1956")
}
# subset the filtered data for these top genes
heatmap_data <- filtered_data[top_50_ids, ]
# get the gene symbols for row names
gene_symbols_heatmap <- top_genes[top_50_ids, "symbol"]

# set the row names of our data to be the gene symbols
final_labels_heatmap <- ifelse(is.na(gene_symbols_heatmap), top_50_ids, gene_symbols_heatmap)
rownames(heatmap_data) <- final_labels_heatmap

# --- 3. THIS IS THE FIX: Order the Columns ---
# Get the sample names in the order you want
untreated_samples <- sample_info$Sample[sample_info$Group == "Untreated"]
osimer_samples <- sample_info$Sample[sample_info$Group == "Osimertinib"]

# Create a single vector of the correct sample order
ordered_sample_names <- c(untreated_samples, osimer_samples)

# Re-order your heatmap data matrix using this vector
heatmap_data_ordered <- heatmap_data_labeled[, ordered_sample_names]

# --- 4. Get your annotation dataframe ---
# (This code is the same as before)
annotation_col <- data.frame(
  Group = sample_info$Group,
  row.names = sample_info$Sample
)

# --- 5. Create the heatmap ---
pheatmap(
  heatmap_data_ordered,         # <-- Use the new ordered matrix
  annotation_col = annotation_col,
  scale = "row",
  cluster_cols = FALSE,         # <-- CRITICAL: Turn off column clustering
  main = "Top 50 DEGs + EGFR (Untreated vs. Osimertinib)"
)

```

## volcano plot

```{r}
## construct a volcano plot to visualize the differential expression results, highlighting the EGFR gene
```{r}
# --- 1. Load required libraries ---
library(ggplot2)
library(ggrepel)

# --- 2. Prepare the data for plotting ---
# Your data is 'top_genes', which is already a data frame.
res_df <- top_genes

# --- 3. Define significance (using FDR < 0.05) ---
# We use the column names from your 'top_genes' table: 'adj.P.Val' and 'logFC'
res_df$significance <- "Not Significant"
res_df$significance[res_df$adj.P.Val < 0.05 & !is.na(res_df$adj.P.Val)] <- "Significant (FDR < 0.05)"
res_df$significance[res_df$adj.P.Val < 0.05 & abs(res_df$logFC) > 1 & !is.na(res_df$adj.P.Val)] <- "Significant (FDR & LFC)"

# --- 4. Create Y-axis variable ---
# We use 'adj.P.Val'. (Note: No y-axis cap is needed, as these p-values are not Inf)
res_df$plot_padj <- -log10(res_df$adj.P.Val)

# --- 5. Create the label for EGFR ---
# This uses the 'symbol' column you created
res_df$gene_label <- ifelse(res_df$symbol == "EGFR", "EGFR", NA)

# --- 6. Create the plot and save it to a variable ---
my_volcano_plot_gse249721 <- ggplot(res_df, aes(x = logFC, y = plot_padj)) +
  
  # Draw all points (Red/Blue/Grey)
  geom_point(aes(color = significance), size = 1) +
  
  # Draw a black point on top of EGFR
  geom_point(data = subset(res_df, symbol == "EGFR"), color = "black", size = 1.5) +
  
  # Add the text label for EGFR
  geom_text_repel(aes(label = gene_label), na.rm = TRUE, box.padding = 0.5) +
  
  # Set the colors
  scale_color_manual(values = c(
    "Not Significant" = "grey",
    "Significant (FDR < 0.05)" = "blue",
    "Significant (FDR & LFC)" = "red"
  )) +
  
  # Set the titles
  labs(
    title = "Volcano Plot: Osimertinib vs. Untreated (GSE249721)",
    subtitle = "limma analysis (significance by FDR < 0.05)",
    x = "log2 Fold Change",
    y = "-log10 (FDR)"
  ) +
  theme_minimal()

# --- 7. Print the plot to your R window/notebook ---
print(my_volcano_plot_gse249721)

# --- 8. Save the plot as a square PNG file ---
# (I've used a file path based on your other script)
ggsave(
  filename = "laboratory_bio_datamining/assessing_mutation_frequency_EGFR_NSCLC/RNA_data_analysis/images/GSE249721_volcano_plot_square.png",
  plot = my_volcano_plot_gse249721,
  width = 8,
  height = 8,
  units = "in",
  dpi = 300
)
```


