
## install libraries 
```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("GEOquery")
BiocManager::install("DESeq2")
BiocManager::install("apeglm")
```
```{r}
library(GEOquery)


library(DESeq2)



library(apeglm)
```


load the data from the file
dataset: GSE278222_HCC827_counts.txt

```{r}
path_GSE <- "/Users/alvise/Documents/uni/semestre3/laboratory_bio_datamining/data/RNA/expression_matrices/GSE278222_HCC827_counts.txt"
# it is a feature count file coming from rnaseq parse it the following way 
full_data <- read.delim(
    file = path_GSE,
    header = TRUE,
    comment.char = "#"
)

# extract the espression matrix columns 7 to the end
expression_matrix <- full_data[ , 7:ncol(full_data)]

#use the first column "Geneid" as row names
rownames(expression_matrix) <- full_data$Geneid

# check the new clean matrix
dim(expression_matrix)
print(expression_matrix[1:6, 1:6])

# save the expression matrix as rds file for future use
saveRDS(expression_matrix, file = "/Users/alvise/Downloads/GSE278222_expression_matrix.rds")
```

## load the data form the supplementary files
```{r}
# download the supplementary files and save them in a specific folder
path_1 <- "/Users/alvise/Documents/uni/semestre3/laboratory_bio_datamining/data/RNA/expression_matrices/GSE278222"
supp_files <- getGEOSuppFiles("GSE278222", baseDir = path_1)
# check the downloaded files
list.files(path_1)
```

# load the rds file
```{r}
matrix_path <- "/Users/alvise/Downloads/GSE278222_expression_matrix.rds"
expression_matrix <- readRDS(file = matrix_path)
head(expression_matrix)
```

## preprocessing the data

create groups based on treatment conditions
the two groups are: "Osimertinib" and "Untreated"
```{r}
# check the column names to identify treatment conditions
colnames(expression_matrix)
# one group is OsiR and the other is HCC827 then there is a dot and a number

# create a new data frame with the sample names and their corresponding groups
sample_info <- data.frame(
    Sample = colnames(expression_matrix),
    Group = ifelse(grepl("OsiR", colnames(expression_matrix)), "Osimertinib", "Untreated")
)   
# check the sample information
head(sample_info)
```

## check the distribution of gene expression in samples in each group
create a boxplot to visualize the distribution of expressionvalue across smaples. 
include the two groups with different colors. 
save image here: laboratory_bio_datamining/assessing_mutation_frequency_EGFR_NSCLC/RNA_data_analysis/images


```{r}
png(filename = "/Users/alvise/Documents/uni/semestre3/laboratory_bio_datamining/assessing_mutation_frequency_EGFR_NSCLC/RNA_data_analysis/images/GSE278222_boxplot_expression_values.png",
    width = 800, height = 600)
boxplot(expression_matrix,
        main = "Gene Expression Distribution in HCC827 Cell Line Samples (GSE278222)",
        xlab = "Samples",
        ylab = "Expression Values",
        col = ifelse(sample_info$Group == "Osimertinib", "lightblue", "lightgreen"),
        las = 2, # rotate x-axis labels for better readability
        outline = FALSE) # remove outliers for better visualization
dev.off()

## print the highes and lowest expression values to check normalization
print(paste("Maximum expression value:", max(expression_matrix)))
print(paste("Minimum expression value:", min(expression_matrix)))
```

The original paper seel link: https://www.nature.com/articles/s41419-025-07711-z#Sec2
The data is NOT normalized or log transformed. values range form 0 to approx 6 million. 

## for this reason we procede with differential expression analysis using deseq2

```{r}
# 1. Create the DESeqDataSet (DDS) object
# This object stores your counts and sample info together
dds <- DESeqDataSetFromMatrix(
  countData = expression_matrix,
  colData = sample_info,
  design = ~ Group  # Tell DESeq2 to model based on the "Group" column
)

# 2. (Optional but Recommended) Pre-filtering
# This removes genes with almost no counts, speeding up the analysis.
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep, ]

# 3. Set the baseline group (CRITICAL)
# We must tell DESeq2 that "Untreated" is the baseline for comparison.
dds$Group <- relevel(dds$Group, ref = "Untreated")

# 4. Run the main DESeq2 analysis
# This one command handles normalization, dispersion estimation, and fitting.
dds <- DESeq(dds)

# 5. Get the results table
# This extracts the comparison: "Osimertinib" vs. "Untreated"
res <- results(dds, name = "Group_Osimertinib_vs_Untreated")

# 6. (Optional but Recommended) Get "shrunken" results for better visualization
# This shrinks the log-fold changes for low-count genes, making them more reliable.
# You may need to install this: BiocManager::install("apeglm")
resLFC <- lfcShrink(dds, coef = "Group_Osimertinib_vs_Untreated", type = "apeglm")

# 7. Check your final results!
# Summary of how many genes are significant:
summary(res)

# View the top genes from the shrunken results (best for plotting):
print("Top significant genes (shrunken LFC):")
print(head(resLFC[order(resLFC$padj), ]))
```

### add gene symbols
clean up the gene ids and convert them to gene symbols and add them to the results table

```{r}
# 1. Install/load the human genome annotation database
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
# BiocManager::install("org.Hs.eg.db")
library(org.Hs.eg.db)

# 2. Get the Ensembl IDs from your DESeq2 results
ensembl_ids_with_version <- rownames(resLFC)

# 3. Clean the IDs (remove the ".9", ".16" etc.)
# This finds the dot "." and everything after it, and replaces it with nothing.
ensembl_ids_cleaned <- sub("\\..*", "", ensembl_ids_with_version)

# 4. Use mapIds to translate the *cleaned* IDs
gene_symbols <- mapIds(org.Hs.eg.db,
                       keys = ensembl_ids_cleaned,
                       column = "SYMBOL",
                       keytype = "ENSEMBL",
                       multiVals = "first")

# 5. Add the symbols as a new column to your results table
# We match the cleaned IDs to the map to re-order them correctly
resLFC$symbol <- gene_symbols[ensembl_ids_cleaned]

# 6. Check your new, readable results!
print("Results with gene symbols:")
print(head(resLFC[order(resLFC$padj), ]))
```

check the expression results for EGFR gene
```{r}
# --- 2. THIS IS THE CORRECT WAY TO LOOK FOR EGFR ---
# Search for the "EGFR" string in the 'symbol' column you just made

# Check if "EGFR" is in your new symbol column
present <- "EGFR" %in% resLFC$symbol
print(paste("Is EGFR in the results?", present))

# If TRUE, print the results row for EGFR
if (present) {
  print("Results for EGFR:")
  # Use 'which' to get the row(s) safely, ignoring NAs
  print(resLFC[which(resLFC$symbol == "EGFR"), ])
}
```
                    baseMean log2FoldChange     lfcSE     pvalue       padj     symbol

ENSG00000146648.20    175610      -0.218491 0.0665007 0.00105102 0.00280721       EGFR

## construct a volcano plot to visualize the differential expression results, highlighting the EGFR gene
```{r}
# --- 1. Load required libraries ---
library(ggplot2)
library(ggrepel)

# --- 2. Prepare the data for plotting ---
# Convert the DESeq2 results object to a data frame
res_df <- as.data.frame(resLFC)

# --- 3. Define significance (using 0.05) ---
res_df$significance <- "Not Significant"
res_df$significance[res_df$padj < 0.05 & !is.na(res_df$padj)] <- "Significant (FDR < 0.05)"
res_df$significance[res_df$padj < 0.05 & abs(res_df$log2FoldChange) > 1 & !is.na(res_df$padj)] <- "Significant (FDR & LFC)"

# --- 4. Set the y-axis cap to 25 ---
y_cap <- 25
res_df$plot_padj <- -log10(res_df$padj)
res_df$plot_padj[is.infinite(res_df$plot_padj)] <- y_cap # Cap 'Inf' p-values
res_df$plot_padj[res_df$plot_padj > y_cap] <- y_cap     # Cap all other high p-values

# --- 5. Create the label for EGFR ---
res_df$gene_label <- ifelse(res_df$symbol == "EGFR", "EGFR", NA)

# --- 6. Create the plot and save it to a variable ---
my_volcano_plot <- ggplot(res_df, aes(x = log2FoldChange, y = plot_padj)) +
  
  # Draw all points (Red/Blue/Grey)
  geom_point(aes(color = significance), size = 1) +
  
  # Draw a black point on top of EGFR
  geom_point(data = subset(res_df, symbol == "EGFR"), color = "black", size = 1.5) +
  
  # Add the text label for EGFR
  geom_text_repel(aes(label = gene_label), na.rm = TRUE, box.padding = 0.5) +
  
  # Set the colors (with corrected 0.05 labels)
  scale_color_manual(values = c(
    "Not Significant" = "grey",
    "Significant (FDR < 0.05)" = "blue",
    "Significant (FDR & LFC)" = "red"
  )) +
  
  # Set the titles (with corrected 0.05 subtitle)
  labs(
    title = "Volcano Plot: Osimertinib vs. Untreated",
    subtitle = "GSE278222 (y-axis capped, significance by FDR < 0.05)",
    x = "log2 Fold Change",
    y = "-log10 (FDR)"
  ) +
  theme_minimal()

# --- 7. Print the plot to your R window/notebook ---
print(my_volcano_plot)

# --- 8. Save the plot as a square PNG file ---
ggsave(
  filename = "/Users/alvise/Documents/uni/semestre3/laboratory_bio_datamining/assessing_mutation_frequency_EGFR_NSCLC/RNA_data_analysis/images/GSE278222_volcano_plot_square.png",
  plot = my_volcano_plot,
  width = 8,
  height = 8,
  units = "in",
  dpi = 300
)
```

## construct a heatmap to visualize the expression of top 50 differentially expressed genes including EGFR

```{r}
# --- 1. Install/load required packages ---
if (!require("pheatmap", quietly = TRUE)) install.packages("pheatmap")
if (!require("RColorBrewer", quietly = TRUE)) install.packages("RColorBrewer")
library(pheatmap)
library(RColorBrewer)

# --- 2. Get normalized expression data ---
# We use vst() (variance-stabilizing transformation) on the dds object
# This normalizes the raw counts in a way that's ideal for visualization.
vst_data <- vst(dds, blind = FALSE)
vst_matrix <- assay(vst_data)

# --- 3. Get the list of genes to plot ---
# Get the Ensembl IDs of the top 50 significant genes (ordered by padj)
top_50_ids <- rownames(resLFC[order(resLFC$padj), ])[1:50]

# Get the Ensembl ID for EGFR (use [1] in case 'which' finds multiple)
egfr_id <- rownames(resLFC[which(resLFC$symbol == "EGFR"), ])[1]

# Combine them into one list and remove any duplicates
genes_to_plot_ids <- union(top_50_ids, egfr_id)

# --- 4. Prepare the Heatmap Data Matrix ---
# Subset the VST matrix to get only the genes we want
heatmap_data <- vst_matrix[genes_to_plot_ids, ]

# Get the gene symbols for these genes from your results
heatmap_labels <- resLFC[genes_to_plot_ids, "symbol"]

# If any symbol is NA, use the Ensembl ID instead
final_labels <- ifelse(is.na(heatmap_labels), genes_to_plot_ids, heatmap_labels)
rownames(heatmap_data) <- final_labels

# --- 5. Order the Matrix Columns (as you requested) ---
# Get sample names for each group
untreated_samples <- sample_info$Sample[sample_info$Group == "Untreated"]
osimer_samples <- sample_info$Sample[sample_info$Group == "Osimertinib"]

# Create a single vector of the correct sample order
ordered_sample_names <- c(untreated_samples, osimer_samples)

# Re-order your heatmap data matrix using this vector
heatmap_data_ordered <- heatmap_data[, ordered_sample_names]

# --- 6. Create the Column Annotation ---
# pheatmap needs a data frame where rownames match the data's column names
annotation_col <- data.frame(
  Group = sample_info$Group,
  row.names = sample_info$Sample
)

# --- 7. Create the Heatmap ---
pheatmap(
  heatmap_data_ordered,         # <-- Use the new ordered matrix
  annotation_col = annotation_col,  # Adds the "Untreated"/"Osimertinib" bar
  scale = "row",                  # CRITICAL: z-scores each gene
  cluster_cols = FALSE,         # CRITICAL: Keeps your "Untreated" -> "Osi" order
  cluster_rows = TRUE,            # Groups similar genes
  
  main = "Top 50 DEGs + EGFR (GSE278222)",
  subtitle = "Expression (VST normalized, row-scaled)",
  
  # Use a nice Blue-White-Red color palette
  color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(100),
  
  # Adjust text size
  fontsize_row = 8
)
```

