
## install libraries 
```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("GEOquery")
library(GEOquery)

```

## load data from GEOquery
dataset: GSE283536
```{r}
data_GEO <- getGEO("GSE283536", GSEMatrix = TRUE)
expression_data <- exprs(data_GEO[[1]])
# check the dimensions of the expression matrix
dim(expression_data)
print(expression_data[0:10, 0:4])
# get metadata
# Extract the sample metadata table
sample_info <- pData(data_GEO[[1]])

# Print the first few columns to see what's inside
# (It has 38 columns, so we just look at the first 6)
print(head(sample_info[, 1:6]))
# Extract and print specific metadata information
sample_info_titles <- sample_info$title
print(sample_info_titles)
# extract at the same time the geo_accession and title columns
sample_metadata <- sample_info[, c("geo_accession", "title")]
print(head(sample_metadata))

```

now that i have a map form the geoa ccessio to the title i can use it to create sample labels for the data. 

```{r}
# check the expression data column names
colnames(expression_data)

# substitute the colnames from the geo_accesison to the title 
print(expression_data[0:5, 0:4])

# Create a named vector to map geo_accession to title
accession_to_title <- setNames(sample_metadata$title, sample_metadata$geo_accession)
# add a row with the titles as colnames
colnames(expression_data) <- accession_to_title[colnames(expression_data)]
#check the updated column names
colnames(expression_data)
# cehch the head of the expression data with new colnames
print(expression_data[0:5, 0:4])
```

now in the column names of the expression matrix i have the titles of the samples instead of the geo_accession.

here i want to dave the expression matrix as a csv file for further analysis.
```{r}
# save the expression data as a rds file
save_rds_path <- "/Users/alvise/Documents/uni/semestre3/laboratory_bio_datamining/data/RNA/GSE283536/expression_data_GSE283536.rds"
saveRDS(expression_data, file = save_rds_path)
```
now load the rds file to check that it was saved correctly
```{r}
# load the expression data from the rds file
loaded_expression_data <- readRDS(save_rds_path)
# check the dimensions of the loaded expression matrix
dim(loaded_expression_data)
```

## create sample groups based on the titles
the two groups are Osimertinib and Control. 
If "CTL" is present in the title then it is a control sample, otherwise it is an Osimertinib sample.

```{r}
# create a data frame with sample names and their corresponding groups
sample_groups <- data.frame(
    sample_name = colnames(loaded_expression_data),
    group = ifelse(grepl("CTL", colnames(loaded_expression_data)), "Control", "Osimertinib")
)
# print the sample groups
print(sample_groups)
```

## check the distribution of gen expression in samples in each group
createa. boxplot ro vosualize distribution across samples. 
include the two gorups with different colors. 
save images at the specified path. 
```{r}
# load ggplot2 for visualization
library(ggplot2)
library(reshape2)
# path to save file
save_box_path <- "/Users/alvise/Documents/uni/semestre3/laboratory_bio_datamining/assessing_mutation_frequency_EGFR_NSCLC/RNA_data_analysis/images"
# create the image
png(filename = file.path(save_box_path, "expression_boxplot_GSE283536.png"), width = 1200, height = 800)
# create boxplot
boxplot(loaded_expression_data,
        main = "Gene Expression Distribution Across Samples (GSE283536)",
        xlab = "Samples",
        ylab = "Expression Levels",
        las = 2,
        col = ifelse(sample_groups$group == "Control", "lightblue", "lightgreen")
        )
dev.off()

# additionally chech the distribution of the expression values
summary(loaded_expression_data)

# --- 1. Apply log2 transformation ---
# We add 1 to avoid log2(0)
log2_data <- log2(loaded_expression_data + 1)

# --- 2. Check the data again ---
# This summary will now look "normal" (e.g., max ~ 10-12)
print(summary(log2_data))

# --- 3. Create a boxplot to visualize the normalized data ---
boxplot(log2_data, 
        main = "Boxplot of Log2-Transformed Data (GSE283536)",
        ylab = "Log2 Expression",
        las = 2) # las=2 rotates sample names for readability
```

values range between 0 and 1500 approximately. This means that i need to log transform the data for further analysis.
already log 2 transformed in the code block above. 

now procede with limma to do differential expression analysis between the two groups.

```{r}
# load limma library
library(limma)
# compute the variance for each gene across all samples
gene_variances <- apply(log2_data, 1, var)
# keep only genes that have variance greater than 0
keep_genes <- gene_variances > 0
filtered_data <- log2_data[keep_genes, ]

# see hom many genes were kept
print(paste("Number of genes kept after filtering:", nrow(filtered_data)))
print(paste("Number of genes removed after filtering:", nrow(log2_data) - nrow(filtered_data)))

# create design matrix for the two groups
design_m <- model.matrix(~ 0 + factor(sample_groups$group))
colnames(design_m) <- c("Control", "Osimertinib")
contrast.matrix <- makeContrasts(Osimertinib - Control, levels = design_m)

# fit the linear model
fit <- lmFit(filtered_data, design_m)
fit2 <- contrasts.fit(fit, contrast.matrix)
fit2 <- eBayes(fit2)

# get the differentially expressed genes
top_genes <- topTable(fit2, adjust.method = "BH", number = Inf)
# view the top differentially expressed genes
head(top_genes)
```

## now translate the gene ids into gene symbols

```{r}
# load the org.Hs.eg.db library for gene annotation
BiocManager::install("org.Hs.eg.db")
library(org.Hs.eg.db)
```

```{r}
# check the gnee names in the top_genes data frame
head(rownames(top_genes))

# extract tje featured data (anotaion table) from my geo object
feature_data <- fData(data_GEO[[1]])
# check the first few rows of the feature data
head(feature_data)
print(colnames(feature_data))
```
here i'm able to identify the column GENE_SYMBOL 
               LOCUSLINK_ID  GENE_SYMBOL
A_19_P00315452    100130938 LOC100130938
A_19_P00315502    102723721 LOC102723721
A_19_P00315506         9674     KIAA0040
A_19_P00315551       400043    LOC400043
A_19_P00315581       400955    LINC01122
A_19_P00315584    100130691 LOC100130691

```{r}
# add the gene symbols to the top_genes data frame
top_genes$GENE_SYMBOL <- feature_data[rownames(top_genes), "GENE_SYMBOL"]
# view the updated top_genes data frame with gene symbols
print(head(top_genes))

# now find EGFR 
egfr_row <- top_genes[which(top_genes$GENE_SYMBOL == "EGFR"), ]
print(egfr_row)
```

now clean up the top_genes data frame by setting the GENE_SYMBOL as rownames and removing the old rownames
- this cannot be done directly because of duplicated gene symbols. 
- for this reason i wan to extract all the gene_ids (rownames) corresponding to EGFR and then subset the top_genes data frame to include only those rows.
```{r}
egfr_rows_inds <- which(top_genes$GENE_SYMBOL == "EGFR")  # these are the row indices for EGFR
print(egfr_rows_inds)
# now extract the row names (gene ids) for these indices
egfr_gene_ids <- rownames(top_genes)[egfr_rows_inds]
print(egfr_gene_ids)
```



## construct a heat map for the top 30 differentially expressed genes
this will include EGFR as well.
```{r}
library(pheatmap)
library(RColorBrewer)
# select the top 30 differentially expressed genes
top30_genes <- rownames(top_genes)[1:30]
# also include EGFR     
top30_genes <- unique(c(top30_genes, egfr_gene_ids))
# subset the filtered data to include only the top 30 genes
heatmap_data <- filtered_data[top30_genes, ]
# get the gene symbols for the top 30 genes
gene_symbols_top30 <- feature_data[top30_genes, "GENE_SYMBOL"]
# set the row names of the heatmap data to gene symbols
rownames(heatmap_data) <- gene_symbols_top30

# create annotation for the columns based on sample groups
annotation_col <- data.frame(
    Group = sample_groups$group
)

# set row names for annotation to match column names of heatmap data
rownames(annotation_col) <- sample_groups$sample_name

# create the heatmap
pheatmap(
  heatmap_data,
  annotation_col = annotation_col,
  main = "Heatmap of Top 30 Differentially Expressed Genes (GSE283536)",
  fontsize_row = 8,
  fontsize_col = 6,
  
  # --- Additions for consistency ---
  scale = "row",          # z-scores the data, which is critical for visualization
  cluster_cols = FALSE,   # Keeps your sample order (e.g., Untreated, then Osimertinib)
  cluster_rows = TRUE,    # Clusters the genes
  color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(100), # Your usual color scheme
  
  # --- This is the corrected part ---
  # filename, width, and height go *inside* the pheatmap() call
  filename = file.path(save_box_path, "heatmap_top30_genes_GSE283536.png"),
  width = 10,
  height = 8
) # <-- The final parenthesis goes here
```

## construct a volcano plot for the differential expression results
include gene labels for EGFR 
```{r}
# --- 1. Load required libraries ---
library(ggplot2)
library(ggrepel)
library(dplyr) # For the 'mutate' function

# --- 2. Prepare the data ---
# Your data is 'top_genes'.
res_df <- top_genes

# --- 3. Define thresholds ---
lfc_threshold <- 1.0  # You had 1, but 1.5 is also common
fdr_threshold <- 0.05

# --- 4. Create the "class" column for coloring ---
res_df <- res_df %>%
  mutate(
    # Clean the GENE_SYMBOL column just in case
    GENE_SYMBOL_CLEAN = trimws(GENE_SYMBOL),
    
    # Create the 'class' column based on logFC and adj.P.Val
    class = case_when(
      logFC > lfc_threshold & adj.P.Val < fdr_threshold ~ "+",
      logFC < -lfc_threshold & adj.P.Val < fdr_threshold ~ "-",
      TRUE ~ "="
    ),
    # Make 'class' a factor to control the legend order
    class = factor(class, levels = c("+", "-", "=")),
    
    # Create the label column. This will label all probes for EGFR.
    gene_label = ifelse(GENE_SYMBOL_CLEAN == "EGFR", "EGFR", NA)
  )

# --- 5. Create the plot (new style) ---
my_volcano_plot_gse283536 <- ggplot(res_df, aes(x = logFC, y = -log10(adj.P.Val))) +
  
  # Draw all points, colored by our new 'class' column
  geom_point(aes(color = class), alpha = 0.7) +
  
  # Add text labels for the genes in 'gene_label' (i.e., all EGFR probes)
  geom_text_repel(aes(label = gene_label),
                  na.rm = TRUE,
                  color = "black", # Make the label black
                  box.padding = 0.5) +
  
  # Apply the colors
  scale_color_manual(values = c("=" = "grey70", "+" = "firebrick3", "-" = "blue3")) +
  
  # Add the dashed threshold lines (using your thresholds)
  geom_vline(xintercept = c(-lfc_threshold, lfc_threshold), linetype = "dashed", color = "grey20") +
  geom_hline(yintercept = -log10(fdr_threshold), linetype = "dashed", color = "grey20") +
  
  # Add the theme
  theme_bw() +
  
  # Add labs
  labs(
    title = "Volcano Plot: Osimertinib vs. Untreated (GSE283536)",
    subtitle = paste("Significance: FDR <", fdr_threshold, "and |logFC| >", lfc_threshold),
    x = "log2 Fold Change",
    y = "-log10 (FDR)",
    color = "Significance" # Adds a title to the legend
  )

# --- 6. Print the plot ---
print(my_volcano_plot_gse283536)

# --- 7. Save the plot ---
ggsave(
  filename = file.path(save_box_path, "volcano_plot_GSE283536.png"),
  plot = my_volcano_plot_gse283536,
  width = 10,
  height = 8
)
```

## export the result of the differential expression analysis containing the genesymbols to a tsv file
```{r}
# define the save path
save_path_12 <- "/Users/alvise/Documents/uni/semestre3/laboratory_bio_datamining/assessing_mutation_frequency_EGFR_NSCLC/top_common_genes"
# save the top_genes data frame as a tsv file
write.table(top_genes, file = file.path(save_path_12, "GSE283536_DE_genes_with_symbols.tsv"), sep = "\t", row.names = TRUE, quote = FALSE)
```

     
