
```{r}
library(GEOquery)
library(Biobase)
library(tidyverse)
library(limma)
```
## download the data from GEO
```{r}
save_path <- "/Users/alvise/Documents/uni/semestre3/laboratory_bio_datamining/data/RNA/expression_matrices/GSE202859"
GSE202859 <- getGEOSuppFiles("GSE202859", makeDirectory = FALSE, baseDir = save_path)
```

## load the file
```{r}
# file is already unzipped
file_path <- "/Users/alvise/Documents/uni/semestre3/laboratory_bio_datamining/data/RNA/expression_matrices/GSE202859/GSE202859_TPMs_allsamples.txt"
# read the data
expression_data <- read.delim(file_path, header = TRUE, row.names = 1)
# check the dimensions of the data
dim(expression_data)
# check the first few rows of the data
head(expression_data)
```

extract only the sample (columns) of the HCC827 cell line
```{r}
# get the column names
colnames(expression_data)
# select only the columns that contain "HCC827"
hcc827_data <- expression_data %>% select(contains("HCC827"))
# check the dimensions of the new data
dim(hcc827_data)
# check the first few rows of the new data
head(hcc827_data)
# check the structure
str(hcc827_data)
```

the data is already in TPMs, so no need to normalize it further. jut compute the log2 transformation
```{r}
# 1. Convert the character data frame to a numeric matrix
# We use 'sapply' to safely convert each column to numbers
hcc827_numeric <- sapply(hcc827_data, as.numeric)

# 2. Restore the row names (Gene IDs)
# sapply returns a matrix but can lose row names, so we put them back
rownames(hcc827_numeric) <- rownames(hcc827_data)

# 3. Handle any NAs introduced during conversion
# (If a cell had text like "Null", it becomes NA. We replace with 0)
hcc827_numeric[is.na(hcc827_numeric)] <- 0

# 4. NOW you can run the log transformation safely
hcc827_log2 <- log2(hcc827_numeric + 1)

# 5. Check the result - it should now show numbers, not "chr"
print("Structure after fix:")
str(hcc827_log2)

print("First rows:")
print(head(hcc827_log2))
```

## save the new data frame
```{r}
# save the data frame as a .rds file in the same directory
saveRDS(hcc827_log2, file = file.path(save_path, "GSE202859_HCC827_TPMs_log2.rds"))
```

## now produce a boxplot of the log2 transformed data
```{r}
# convert the matrix to a data frame for ggplot
hcc827_log2_df <- as.data.frame(hcc827_log2)
# add a gene column for ggplot
hcc827_log2_df$Gene <- rownames(hcc827_log2_df)
# reshape the data to long format
hcc827_long <- hcc827_log2_df %>% pivot_longer(cols = -Gene, names_to = "Sample", values_to = "Log2_TPM")
# create the boxplot
ggplot(hcc827_long, aes(x = Sample, y = Log2_TPM)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

from the boxplot we can see that the data is well normalized across samples.

## procede with differential expression analysis using limma

first produce the design matrix with the conditions of interest:
- pre treatment
- post treatment 

```{r}
## procede with dea using limma

# 1. Define the design and sample info safely
# Ensure Condition is a factor with "Pre_Treatment" as the first (reference) level
# This prevents the column order mix-up
sample_info$Condition <- factor(sample_info$Condition, levels = c("Pre_Treatment", "Post_Treatment"))

# Create the design matrix
design <- model.matrix(~ 0 + Condition, data = sample_info)

# Rename columns safely (now we know the order is Pre, then Post)
colnames(design) <- c("Pre_Treatment", "Post_Treatment")
print("Design Matrix:")
print(design)

# 2. FILTERING STEP (Crucial for preventing eBayes errors)
# Ensure data is numeric matrix
expr_matrix <- as.matrix(hcc827_log2)

# Calculate variance and expression stats
gene_variances <- apply(expr_matrix, 1, var)
# Check if gene is expressed (Log2TPM > 1) in at least 2 samples
is_expressed <- rowSums(expr_matrix > 1) >= 2

# Define filter: Variance > 0 AND Expressed in at least 2 samples
keep <- (gene_variances > 0) & is_expressed
filtered_data <- expr_matrix[keep, ]

print(paste("Original genes:", nrow(expr_matrix)))
print(paste("Filtered genes:", nrow(filtered_data)))

# 3. Run Limma on Filtered Data
# fit the linear model
fit <- lmFit(filtered_data, design)

# create the contrast matrix
contrast.matrix <- makeContrasts(Post_vs_Pre = Post_Treatment - Pre_Treatment, levels = design)

# fit the contrasts
fit2 <- contrasts.fit(fit, contrast.matrix)

# apply empirical Bayes moderation
fit2 <- eBayes(fit2)

# get the top differentially expressed genes
topTable_results <- topTable(fit2, adjust.method = "BH", number = Inf)

# view the results
head(topTable_results)
```

## now extract the gene names and put them in a new column called "gene_sym"

at the moment the row names of the data contain a string with both Gene id and gene symbol separated by a "|". 
now take the row names and split the string at the "|" character and extract both the gene is 
and the gene symbol and put them in two new columns called "gene_id" and "gene_sym"
```{r}
library(dplyr)
library(tidyr)
library(tibble)

# 1. Convert rownames to a column, then split
topTable_results <- topTable_results %>%
  tibble::rownames_to_column(var = "Gene_Info") %>%
  separate(Gene_Info, 
           into = c("gene_id", "gene_sym"), 
           sep = "\\|", 
           extra = "merge", # If there are extra parts, merge them into the last column
           fill = "right")  # If parts are missing, fill with NA on the right

# 2. (Optional) Clean up any whitespace
topTable_results$gene_sym <- trimws(topTable_results$gene_sym)
topTable_results$gene_id <- trimws(topTable_results$gene_id)

# 3. Check the results
head(topTable_results)
```

now extract the results for EGFR, where in the gene_sym column i find "EGFR"

```{r}
egfr_results <- topTable_results %>% filter(gene_sym == "EGFR")
# view the results for EGFR
print(egfr_results)
```

          gene_id gene_sym     logFC  AveExpr        t      P.Value      adj.P.Val        B
1 ENSG00000146648     EGFR -2.548317 9.075807 -15.2799 2.724419e-06   7.111548e-05 5.310147


## heatmap of the top 30 differentially expressed genes including EGFR in nor present already

```{r}
library(pheatmap)
library(RColorBrewer)
library(dplyr)
library(tibble)

# 1. get top 30 by adjusted p-value and ensure EGFR included if present in results
top_genes_tbl <- topTable_results %>% arrange(adj.P.Val) %>% slice_head(n = 30)

egfr_row <- topTable_results %>% filter(gene_sym == "EGFR")
if(nrow(egfr_row) > 0 && ! any(top_genes_tbl$gene_id == egfr_row$gene_id)){
  top_genes_tbl <- bind_rows(top_genes_tbl, egfr_row) %>% distinct(gene_id, .keep_all = TRUE)
}

selected_ids <- top_genes_tbl$gene_id

# 2. map expression rows to gene_id (text before '|') and subset filtered_data
rownames_gene_id <- sub("\\|.*", "", rownames(filtered_data))
keep_idx <- which(rownames_gene_id %in% selected_ids)
expr_subset <- filtered_data[keep_idx, , drop = FALSE]

# 3. set rownames to gene symbols when available (fallback to gene_id)
id2sym <- topTable_results %>% select(gene_id, gene_sym) %>% distinct(gene_id, .keep_all = TRUE)
sym_map <- setNames(ifelse(is.na(id2sym$gene_sym) | id2sym$gene_sym == "", id2sym$gene_id, id2sym$gene_sym), id2sym$gene_id)
rownames(expr_subset) <- sym_map[sub("\\|.*", "", rownames(expr_subset))]

# 4. scale rows (z-score) for visualization
expr_scaled <- t(scale(t(expr_subset)))

# 5. prepare sample annotation if sample_info available and matches columns
annotation_col <- NULL
if(exists("sample_info")){
  ann <- sample_info
  if("Sample" %in% colnames(ann)){
    ann <- ann %>% column_to_rownames("Sample")
  }
  if(all(colnames(expr_scaled) %in% rownames(ann))){
    annotation_col <- ann[colnames(expr_scaled), , drop = FALSE]
  }
}

# 6. draw heatmap
my_palette <- colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(100)

pheatmap(expr_scaled,
         color = my_palette,
         cluster_rows = TRUE,
         cluster_cols = FALSE,
         show_rownames = TRUE,
         show_colnames = TRUE,
         fontsize_row = 8,
         fontsize_col = 8,
         annotation_col = annotation_col,
         main = "Top DE genes (incl. EGFR) — scaled TPM (log2)")
```

## volcanoplot of the differential expression results
point out EGFR in the plot

```{r}
library(ggplot2)
# ggrepel used for non-overlapping labels if available
# dplyr is already loaded earlier; use it here
volc_df <- topTable_results %>%
  mutate(logP = -log10(P.Value),
         signif = (adj.P.Val < 0.05 & abs(logFC) > 1))

# Choose labels: always include EGFR if present, plus up to 4 top significant genes
label_genes <- volc_df %>% filter(gene_sym == "EGFR")
if(nrow(label_genes) == 0){
  label_genes <- volc_df %>% filter(signif) %>% arrange(adj.P.Val) %>% slice_head(n = 5)
} else {
  extra <- volc_df %>% filter(signif, gene_sym != "EGFR") %>% arrange(adj.P.Val) %>% slice_head(n = 4)
  label_genes <- bind_rows(label_genes, extra) %>% distinct(gene_sym, .keep_all = TRUE)
}

# Basic volcano plot
p <- ggplot(volc_df, aes(x = logFC, y = logP)) +
  geom_point(aes(color = signif), alpha = 0.6, size = 1.6) +
  scale_color_manual(values = c("FALSE" = "grey60", "TRUE" = "red")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "blue", size = 0.3) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "blue", size = 0.3) +
  xlab("log2 Fold Change") + ylab("-log10(P-value)") +
  ggtitle("Volcano plot — Post vs Pre") +
  theme_minimal() +
  theme(legend.position = "none")

# Add labels: prefer ggrepel if available, otherwise geom_text
if(requireNamespace("ggrepel", quietly = TRUE)){
  p <- p + ggrepel::geom_text_repel(data = label_genes, aes(label = gene_sym),
                                    size = 3, max.overlaps = 20)
} else {
  p <- p + geom_text(data = label_genes, aes(label = gene_sym),
                     vjust = -0.5, size = 3)
}

print(p)
```

## for the meta analyisis i will save the results as a .csv file
in particular save the all the differentially expressed genes in a .tsv file


```{r}
output_dir <- "/Users/alvise/Documents/uni/semestre3/laboratory_bio_datamining/assessing_mutation_frequency_EGFR_NSCLC/top_common_genes"
# select top 1000 genes by adjusted p-value
top_results <- topTable_results
write.table(top_results, file = file.path(output_dir, "GSE202859_DE_results_with_symbols.tsv"), sep = "\t", row.names = FALSE)
```

## 